State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2021-03-16 analysis/paper/paper.Rmd
made some preliminary revisions including beefing up the methods and others

#------------------------------------------------------
# Re-format res_df into long format of point ests.,
# CI's and CIin for both Models : STADEM, SCOBI
#------------------------------------------------------
res_long_df <- res_df %>%
  pivot_longer(cols = -(Scenario:Truth),
               names_to = "key",
               values_to = "point") %>%
  separate(key, into = c("Model","est"), sep = "_") %>%
  pivot_wider(names_from = "est",
              values_from = "point") %>%
  mutate(Model = recode(Model, 
                        "ISEMP" = "STADEM")) %>%
  rowwise() %>%
  mutate(se = (uppCI - lowCI) / (2 * qnorm(0.975)),
         cv = se / est,
         # bias = est - Truth,
         bias = Truth - est,
         rel_bias = bias / Truth,
         MSE = mean((bias)^2),
         inCI = ifelse(inCI == 1, T, ifelse(inCI == 0, F, NA)),
         Window = ifelse(grepl("H",Scenario),"Window_Err_High",
                         ifelse(grepl("L",Scenario),"Window_Err_Low","No Window Error")),
         Night = ifelse(grepl("N-R",Scenario),"Yes","No"))
```

@@

A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both sp/sum Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group included all tags (both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. The proportion of tags detected leaving the passage system that were also caught in the trap each week was assumed to reflect the trap rate experienced by target species as they migrated through the ladder. The sample size of previously tagged fish detected at Lower Granite Dam influences the uncertainty in that trap rate, which is also informed by the number of adults caught in the trap and the window counts. The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others. Therefore, we use the mark-recapture approach to estimate a "true" trapping rate.

We used PIT tag observation data from the adult fish ladder to estimate both nighttime passage and re-ascension rates; each estimated on a weekly basis. The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder. The re-ascension rate incorporated the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time and the total number of tags leaving the fish ladder. Previously, we looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for estimated nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@Royle2008]. Our model is implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM adjusted the window counts for the nighttime passage rate, and both window and trap estimates for the re-ascension rate to estimate the number of unique fish that crossed the dam. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

We tested the STADEM model on a variety of simulated data sets. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish was given a date of ladder ascension, based on its population and the range of observed run timing for that population. Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be "caught" in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs. Further details about simulation procedures can be found in [Appendix B](#append2).

We grouped the simulation results by fish origin and evaluated them by several measures. Relative bias is the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth". We estimated the precision by examining the average coefficient of variation (CV) of the estimates. We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate. Finally, we evaluated coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.

@@

STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)). Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%. The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks. The coverage probabilities always exceeded 95% across all scenarios. The RMSE was near 500 for each scenario, representing an estimate within 2% of the true value, demonstrating the accuracy of STADEM (Table \@ref(tab:summ-tab)).

@@

We applied STADEM to data from Lower Granite Dam for sp/sum Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Estimates of total unique fish escaping past Lower Granite Dam were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years. CVs ranged from 2.5-7.1% for wild fish, 2.3-5.1% for clipped hatchery fish, 3.4-8.8% for hatchery no-clip fish and 2.2-4.7% for total unique fish past Lower Granite Dam.

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig). Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])). STADEM point estimates were often between estimates based on window counts and those based on the number of fish caught in the adult trap. However, for weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig). That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August. The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.

@@

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder. Our model explicitly accounts for nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years. With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar). Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty. Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses.

@@

@@

```{r lgr-sim-tab}
crossing(`Trap rate` = c('0.15', 
                         "0.15 and 0.00 3 weeks"),
         `Fallback & Re-ascension rate` = c(0.06,
                             0.10),
         # `Re-ascension rate` = 1,
         `Window count error` = factor(c("No Error",
                                         "5% Error",
                                         "10% Error"),
                                       levels = c("No Error",
                                                  "5% Error",
                                                  "10% Error"))) %>%
  mutate(`Nighttime passage rate` = if_else(`Fallback & Re-ascension rate` == 0.06,
                                            0.06,
                                            0.05)) %>%
  mutate(Scenario = paste(rep(c('Baseline',
                                "N-R",
                                "Trap down",
                                "N-R trap down"),
                              each = 3)),
         Scenario = if_else(`Window count error` == "5% Error",
                            paste(Scenario, "Err L"),
                            if_else(`Window count error` == "10% Error",
                                    paste(Scenario, "Err H"),
                                    Scenario))) %>%
  select(Scenario, 
         `Trap rate`:`Fallback & Re-ascension rate`,
         `Nighttime passage rate`,
         `Window count error`) %>%
  kable(digits = 2,
        booktabs = T,
        linesep = "",
        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position")) %>%
  column_spec(column = 2,
              width = "1in") %>%
  column_spec(column = c(3, 4, 5),
              width = "1.2in")

@@

sum_tab %>%
  kable(digits = c(0, 3, 1, 0, 3, 0, 3),
        booktabs = T,
        linesep = "",
        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for spring/summer-run Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates. The dark gray ribbon represents the 95% credible intervals for STADEM estimates, while the light gray ribbon represents the 95% confidence intervals for the trap estimates."}

# get plotting data from particular year
plot_df = tibble(Species = "Chinook",
                 Year = "2014") %>%
  mutate(plot_title = paste(Species, Year)) %>%
  rowwise() %>%
  mutate(plot_data = map2(Species, 
                          Year,
                          .f = function(x, y) {
                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', x, '_', y, '.rda'))
                            # plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
                            plot_df = stadem_mod$summary[grep('^X.new', rownames(stadem_mod$summary)),] %>%  
                              as_tibble(rownames = 'var') %>%
                              mutate(week = as.integer(str_extract(var, "[0-9]+")),
                                     param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
                                     param = recode(param, "tot" = "all")) %>%
                                     # param = ifelse(param == '', 'all', param)) %>%
                              select(var, param, week, everything()) %>%
                              left_join(stadem_list$weeklyData %>%
                                          rename(week = week_num) %>%
                                          rowwise() %>%
                                          mutate(trap_lowCI = max(0, trap_est + trap_est_se * qnorm(0.025)),
                                                 trap_uppCI = trap_est + trap_est_se * qnorm(0.975)) %>%
                                          ungroup())%>%
                              mutate_at(vars(trap_est),
                                        list(~ if_else(. == Inf,
                                                       as.numeric(NA),
                                                       .))) %>%
                              select(-Species)
                            return(plot_df)
                          })) %>%
  unnest(cols = plot_data) %>%
  select(Species, Year, plot_title, everything()) %>%
  filter(param == "all")

@@

```{r child = 'appendices/Appendix_STADEM_descrip.Rmd'}
```

>>>>

#------------------------------------------------------
# Re-format res_df into long format of point ests.,
# CI's and CIin for both Models : STADEM, SCOBI
#------------------------------------------------------
res_long_df <- res_df %>%
  pivot_longer(cols = -(Scenario:Truth),
               names_to = "key",
               values_to = "point") %>%
  separate(key, into = c("Model","est"), sep = "_") %>%
  pivot_wider(names_from = "est",
              values_from = "point") %>%
  mutate(Model = recode(Model, 
                        "ISEMP" = "STADEM")) %>%
  rowwise() %>%
  mutate(se = (uppCI - lowCI) / (2 * qnorm(0.975)),
         cv = se / est,
         bias = est - Truth,
         rel_bias = bias / Truth,
         MSE = mean((bias)^2),
         inCI = ifelse(inCI == 1, T, ifelse(inCI == 0, F, NA)),
         Window = ifelse(grepl("H",Scenario),"Window_Err_High",
                         ifelse(grepl("L",Scenario),"Window_Err_Low","No Window Error")),
         Night = ifelse(grepl("N-R",Scenario),"Yes","No"))
```

@@

<!-- A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both sp/sum Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group included all tags (both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. The proportion of tags detected leaving the passage system that were also caught in the trap each week was assumed to reflect the trap rate experienced by target species as they migrated through the ladder. The sample size of previously tagged fish detected at Lower Granite Dam influences the uncertainty in that trap rate, which is also informed by the number of adults caught in the trap and the window counts. The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others. Therefore, we use the mark-recapture approach to estimate a "true" trapping rate. -->

<!-- We used PIT tag observation data from the adult fish ladder to estimate both nighttime passage and re-ascension rates; each estimated on a weekly basis. The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder. The re-ascension rate incorporated the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time and the total number of tags leaving the fish ladder. Previously, we looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes. -->

@@

### Day-time Passage and Re-ascension Rates

There are two processes that must be accounted for to correct any bias in the window counts: first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted. Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.

The proportion of fish passing the window during non-operational hours, nighttime passage rate, is just the complement of the rate of fish passing during the day when the window is operating. The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].

$$
\begin{aligned}
  y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
  \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
  g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
\end{aligned}
$$

The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fallback and re-ascend the dam. These fish are potentially double-counted at the window, and have the potential to be caught in the fish trap more than once. The number of tags known to be re-ascending the dam each week, $y^{reasc}_t$, is modeled as a binomial process based on the estimated re-ascension rate, $\eta_t$, and the total number of tags crossing the dam that week, $N_t$. The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage [@Shumway2010].

$$
\begin{aligned}
  y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
  \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
  f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
\end{aligned}
$$

We looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

### Total and Weekly Escapement

Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@Royle2008] with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week. We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified. The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].

$$
\begin{aligned}
  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
  e_t &\sim \mathcal{N}(0, \sigma^2_X)
\end{aligned}
$$

The number of fish caught in the trap, $Y^T_t$, for week $t$ is modeled as a binomial process based on the unknown true trap rate that week, $\nu_t$, and the unknown true number of fish crossing the dam that week, $X_t$. The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others. Therefore, we use a mark-recapture approach to estimate a "true" trapping rate. The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002]. The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the the "marked" fish). The proportion of previously marked fish that are caught in the adult trap, $m/M$ can be modeled with a binomial distribution using the same trap rate parameter, $\nu_t$. Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences. The more tagged fish crossing the dam in a particular week, the more certain we can be of the true trap rate.

$$
\begin{aligned}
  m_t &\sim \text{Bin} \left( \nu_t, M_t \right) \\
  Y^T_t &\sim \text{Bin} \left( \nu_t, X_t \right) \\
\end{aligned}
$$

The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open. This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$. In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter. If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.

$$
\begin{aligned}
  X_t^{day} &= X_t * \theta_t \\
  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
\end{aligned}
$$

Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data sources: window counts and fish sampled in the trap. The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week. The use of both estimates allows for an estimation of observer error in the window counts which has previously been inestimable. For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.

### Origin Proportions

After estimating the total number of fish to have crossed Lower Granite each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery clipped fish, $X_{hc,t}$ and hatchery no-clip fish, $X_{hnc, t}$. This is done by estimating a weekly origin proportion vector, $\omega_t$ based on the random sample of fish caught in trap that week, $Y^T_t$. The observed number of wild, $Y^T_{w,t}$, hatchery clipped, $Y^T_{hc,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$. The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish, $\omega_{hc,t}$ is modeled as a random walk, so it can change through time, while smoothing over large weekly flucuations in the proportions due to small sample size. This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.

$$
\begin{aligned}
  \left( Y^T_{w,t}, Y^T_{hc,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
  \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
  \phi_{hc,t} &= 0 \\
  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{hc,t}} \right) \\
  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{hc,t}} \right) \\
  \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
  \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
  d_{w,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
  d_{hnc,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
\end{aligned}
$$

Finally, the number of unique fish by origin crossing Lower Granite each week ($X_{w,t}$, $X_{hc,t}$, and $X_{hnc,t}$) is the product of the total fish crossing that week, $X_t$, multiplied by one minus the re-ascension rate, $(1 - \eta_t)$, and the origin proportion vector, $\omega_t$.

$$
\begin{aligned}
  \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
\end{aligned}
$$

### Model Fitting

The model was fit using the JAGS program [@R-rjags], run with R software [@R-Core-Team2020]. The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference. Variance parameters $\sigma_X, \sigma_\eta, \sigma_\theta$, and $\sigma_\omega$, as well as the initial abundance, $X_1$, and the overdispersion parameter of the negative binomial, $r$, were given half-Cauchy priors with mean of 0 and scale of 100. The initial day-time passage and re-ascension rates, $\theta_1$ and $\eta_1$ were given Uniform(0,1) priors. Finally, $\phi_{w,1}$ and $\phi_{hnc,1}$ were given priors of Uniform(-3,3), in an effort to make $\omega_1$ as uniformative as possible.

@@

We tested the STADEM model on a variety of simulated data sets, constructed to mimic the dynamics at Lower Granite Dam. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish was given a date of ladder ascension, based on its population and the range of observed run timing for that population. Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be "caught" in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and offsetting, and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs. Further details about simulation procedures can be found in [Appendix A](#append1).

We grouped the simulation results by fish origin and evaluated them by several measures. Relative bias is the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth". We estimated the precision by examining the average coefficient of variation (CV) of the estimates. We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias. Finally, we evaluated coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value. We also compared the model estimates of total unique fish with the simulated truth and the window counts, as an evaluation of how well window counts captured this metric.

@@

STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios (Figure \@ref(fig:rel-bias-lgd)). For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R). In those scenarios, they were biased by nearly 5%, the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)). The variance of the window counts also grew as the simulated observer error grew (no error vs. Err_L vs. Err_H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.

Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%. The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks. The coverage probabilities always exceeded 95% across all scenarios. The RMSE was near 500 for each scenario, representing an estimate within 2% of the true value, demonstrating the accuracy and precision of STADEM (Table \@ref(tab:summ-tab)).

@@

We applied STADEM to data from Lower Granite Dam for sp/sum Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Estimates of total unique fish escaping past Lower Granite Dam were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years. Window count bias ranged from -3% to 11% for sp/sum Chinook Salmon and -9.9% to 6.6% for steelhead. Coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past Lower Granite Dam.

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig). Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).

@@

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder. Our model explicitly accounts for nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years. Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty. Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses.

STADEM estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension. STADEM estimates were higher than the window counts when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate. For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig). That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August. The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.

STADEM is an improvement on window counts alone for several reasons. The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects the bias in window counts. That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to (Table \@ref(tab:lgr-tab)). Whether that amount of potential bias is of concern to management agencies is a decision for them, but we have presented a tool to correct that bias. The other justification for STADEM over window counts is that it provides an estimate of observer error for the window counts, leading to estimates of precision for unique fish which are unobtainable with window counts alone. At the same time, STADEM's precision is still quite good, with CV's near 2-3%.  @Hatch1994

@@

With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).

@@

```{r lgr-sim-tab}
crossing(`Trap rate` = c('0.15', 
                         "0.15 and 0.00 3 weeks"),
         `Fallback & Re-ascension rate` = c(0.06,
                             0.10),
         # `Re-ascension rate` = 1,
         `Window count error` = factor(c("No Error",
                                         "5% Error",
                                         "10% Error"),
                                       levels = c("No Error",
                                                  "5% Error",
                                                  "10% Error"))) %>%
  mutate(`Nighttime passage rate` = if_else(`Fallback & Re-ascension rate` == 0.06,
                                            0.06,
                                            0.05)) %>%
  mutate(Scenario = paste(rep(c('Baseline',
                                "N-R",
                                "Trap down",
                                "N-R trap down"),
                              each = 3)),
         Scenario = if_else(`Window count error` == "5% Error",
                            paste(Scenario, "Err L"),
                            if_else(`Window count error` == "10% Error",
                                    paste(Scenario, "Err H"),
                                    Scenario))) %>%
  select(Scenario, 
         `Trap rate`:`Fallback & Re-ascension rate`,
         `Nighttime passage rate`,
         `Window count error`) %>%
  kable(digits = 2,
        booktabs = T,
        linesep = "",
        caption = "Summary of different simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position")) %>%
  column_spec(column = 2,
              width = "1in") %>%
  column_spec(column = c(3, 4, 5),
              width = "1.2in")

@@

sum_tab %>%
  kable(digits = c(0, 3, 1, 0, 3, 0, 3),
        booktabs = T,
        linesep = "",
        caption = "Summary statistics of unique wild fish estimates, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of escapement for spring/summer-run Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates of unique fish. The dark gray ribbon represents the 95% credible intervals for STADEM estimates, while the light gray ribbon represents the 95% confidence intervals for the trap estimates."}

# get plotting data from particular year
plot_df = tibble(Species = "Chinook",
                 Year = "2014") %>%
  mutate(plot_title = paste(Species, Year)) %>%
  rowwise() %>%
  mutate(plot_data = map2(Species, 
                          Year,
                          .f = function(x, y) {
                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', x, '_', y, '.rda'))
                            # plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
                            plot_df = stadem_mod$summary[grep('^X.new', rownames(stadem_mod$summary)),] %>%
                              as_tibble(rownames = 'var') %>%
                              mutate(week = as.integer(str_extract(var, "[0-9]+")),
                                     param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
                                     param = ifelse(param == '', 'all', param),
                                     param = recode(param, 
                                                    "tot" = "all")) %>%
                              select(var, param, week, everything()) %>%
                              left_join(stadem_list$weeklyData %>%
                                          rename(week = week_num) %>%
                                          rowwise() %>%
                                          mutate(trap_lowCI = max(0, trap_est + trap_est_se * qnorm(0.025)),
                                                 trap_uppCI = trap_est + trap_est_se * qnorm(0.975)) %>%
                                          ungroup())%>%
                              mutate_at(vars(trap_est),
                                        list(~ if_else(. == Inf,
                                                       as.numeric(NA),
                                                       .))) %>%
                              select(-Species)
                            return(plot_df)
                          })) %>%
  unnest(cols = plot_data) %>%
  select(Species, Year, plot_title, everything()) %>%
  filter(param == "all")

@@
